---
title: "INF1344 Final Assignment"
author: "Group 10: Athavan Vasantharaj, Franklin Li, Nicholas Field, Tracy Pia, Tsitsi Macherera"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("stringi")
library(kableExtra)
library(tidyverse)
library(moderndive)
library(infer)
library(stringi)
library(car)
```

```{r}
# loading data
df = read.csv("players2022-2023.csv")
df_heights = read.csv("all_seasons.csv")

```


```{r}
# Removing outliers
# Filtering the dataset to only include players that have shot 100 3-point shot attempts and played 800 minutes

df_filtered <- df %>% 
  filter(X3PA > 100  & MP > 800 )

```




```{r}
# join with player height data

# first, clean data so we only use the 2022-2023 season
df_heights <- df_heights %>%
  filter(season == "2022-23")

# then, change non-ASCII characters in player names in df_filtered to ASCII
# using the "stringi" package
df_filtered$Player <- stri_trans_general(df_filtered$Player, "Latin-ASCII")

# then, correct a small number of players whose names are spelled differently
df_filtered$Player <- replace(df_filtered$Player, df_filtered$Player == "Kevin Knox", "Kevin Knox II")
df_filtered$Player <- replace(df_filtered$Player, df_filtered$Player == "Marcus Morris", "Marcus Morris Sr.")
df_filtered$Player <- replace(df_filtered$Player, df_filtered$Player == "OG Anunoby", "O.G. Anunoby")
df_filtered$Player <- replace(df_filtered$Player, df_filtered$Player == "KJ Martin", "Kenyon Martin Jr.")

# finally, join both dataframes on player name
df_joined <- df_filtered %>% 
    inner_join(df_heights, by = c("Player" = "player_name"))

```


```{r}
# create correlation matrix

# create subset df with only numeric variables
correlation_columns <- df_joined %>%
  select(
    "age", 
    "G", # total games played
    "GS", # "games started", with better players usually starting more often,
    "ORB", # offensive rebound
    "MP", # "minutes played"
    "FG.", # "field goals percentage", successes/total shorts, includes both 2-point and 3-point shots
    "X3P.", # "3-point shot percentage"
    "X2P.", # "2-point shot percentage"
    "player_height",
    "player_weight",
    "ts_pct", # "true shooting percentage", an advanced emtric calculating overall shooting ability
    "usg_pct", # "usage percentage", how often a player is used in a team
    "TOV", # turnover rate, i.e., how often the ball is stolen from a player
    "FT." # free-throw percentage
  )

# Section 1: create a correlation matrix and remove the upper triangle
# get correlation matrix via cor()
cor_matrix <- cor(correlation_columns)

# mask the upper triangle and replace NA with space
lower_triangle_with_space <- function(matrix) {
  matrix[upper.tri(matrix)] <- NA
  return(matrix)
}

# Apply the function to the correlation matrix
cor_matrix_lower <- lower_triangle_with_space(cor_matrix)

# Format the correlation matrix to 3 digits
cor_matrix_lower <- format(round(cor_matrix_lower, 3), nsmall = 3)

# Section 2: add columns for mean and standard deviation
# create new matrix with mean and sd columns
df_corr <- correlation_columns

# calculate mean and std dev
means <- colMeans(df_corr)
sds <- apply(df_corr, 2, sd)

# removing the upper triangle of the correlation table
#df_corr <- replace_upper_triangle(df_corr, by = "", diagonal = FALSE)

# Combine mean and standard deviation columns with correlation matrix
result <- cbind(
  Mean = format(round(means, 3), nsmall = 3),
  SD = format(round(sds, 3), nsmall = 3),
  cor_matrix_lower
)

# Show the combined result
knitr::kable(result)

```


```{r}
# change data to long format
correlation_columns_long <- correlation_columns %>%
  pivot_longer(cols = ! c(X3P.), names_to = "variable", values_to = "value")

# create histograms to show distribution of each numeric variable
ggplot(correlation_columns_long, aes(x=value)) + 
  geom_histogram(fill = "red", color = "black") + 
  facet_wrap(~ variable, scale = "free") +
  labs(title = "Histograms of the variables of the NBA 2022-2023 Season dataset", x = "Value", y = "frequency")

```

```{r}
# Creating the scatter plot for each independent variable
ggplot(correlation_columns_long, aes(x=value, y = `X3P.`)) + 
  geom_point(color = "black") +
  facet_wrap(~ variable, scale = "free") +
  labs(title = "Scatter plot of independent variables", x = "Value", y = "3-point %")
```

**Preliminary multilinear model, as baseline, with all possible independent variables**

```{r}
# generate multiple linear regression model

# creating the multiple linear regression model
multiple_regression_model <- lm(formula = X3P. ~ age + G+ GS + ORB + MP + player_height + player_weight + ts_pct + FG. + X2P. + usg_pct + TOV + FT., data = df_joined)

```



```{r}
# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model)

# Checking multicollinearity 
vif_values <- vif(multiple_regression_model)
print(vif_values)

summary(multiple_regression_model)

```
```{r}
# Checking multicollinearity 
vif_values <- vif(multiple_regression_model)
print(vif_values)

```

## Result of team members attempting their own models

**Nick's Model**
```{r}

# creating the multiple linear regression model
multiple_regression_model_nick <- lm(formula = X3P. ~ player_height + ts_pct  + X2P. + FG., data = df_joined)

# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model_nick)


summary(multiple_regression_model_nick)

```

```{r}
# Checking multicollinearity 
vif_values_nick <- vif(multiple_regression_model_nick)
print(vif_values_nick)
```

**Athavan's Model**
```{r}

# creating the multiple linear regression model
multiple_regression_model_athavan <- lm(formula = X3P. ~ age + player_height + player_weight + ts_pct + MP + usg_pct + X2P., data = df_joined)

# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model_athavan)


summary(multiple_regression_model_athavan)

```

```{r}
# Checking multicollinearity 
vif_values_athavan <- vif(multiple_regression_model_athavan)
print(vif_values_athavan)
```


**Combined Best Model (aka Tracy and Franklin's Model v5)**
```{r}

# creating the multiple linear regression model
multiple_regression_model_tracy <- lm(formula = X3P. ~ player_height + ts_pct + usg_pct + ORB + X2P. + TOV + FT., data = df_joined)

# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model_tracy)


summary(multiple_regression_model_tracy)

```

```{r}
# Checking multicollinearity 
vif_values_tracy <- vif(multiple_regression_model_tracy)

knitr::kable(vif_values_tracy)
```

As a side-note, Franklin observed that replacing "True shooting percentage" in the model above with "Effective Field Goals" noticeably improved the adjusted R-squared (up to 0.8124), but tests for multicollinearity led us to reject this variable in our model.
```{r}
multiple_regression_model_efg <- lm(formula = X3P. ~ ORB + player_height + TOV + FT. + eFG. +usg_pct + X2P., data = df_joined) 
```

```{r}
par(mfrow = c(2, 2))
plot(multiple_regression_model_efg)
summary(multiple_regression_model_efg)
```

**Franklin's Model**
```{r}

# creating the multiple linear regression model
multiple_regression_model_franklin <- lm(formula = X3P. ~ age + G + ORB  + MP + player_height + player_weight + ts_pct + usg_pct + X2P., data = df_joined)

# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model_franklin)


summary(multiple_regression_model_franklin)

```


```{r}
# Checking multicollinearity 
vif_values_franklin <- vif(multiple_regression_model_franklin)
print(vif_values_franklin)
```
We originally included "Field Goal percentage (FG.)", but since its VIF was just barely above 5, we removed it. The overall effect on adjusted R-squared was minimal (a slight drop from 0.6908 to 0.69).


**Tsitsi's Model**
```{r}

# creating the multiple linear regression model
multiple_regression_model_tsitsi <- lm(formula = X3P. ~ age + ORB + player_height + ts_pct + G + X2P., data = df_joined)

# Checking the assumptions
par(mfrow = c(2, 2))
plot(multiple_regression_model_tsitsi)


summary(multiple_regression_model_tsitsi)

```

```{r}
# Checking multicollinearity 
vif_values_tsitsi <- vif(multiple_regression_model_tsitsi)
print(vif_values_tsitsi)
```

```{r}
fitted_values <- fitted(multiple_regression_model_tracy)

ggplot(multiple_regression_model_tracy, aes(x = df_joined$X3P., y = fitted_values)) +

  geom_point() + geom_smooth(method = 'lm') + labs(title = "Observed vs fitted/predicted", x = "Observed", y = "fitted")

```
